### 2015-10-07 untitled.tex
edited untitled.tex

Theorem 5 in \cite{Raj:suspension1991} provides a very positive result for the analysis of deferrable task sets subject to period enforcement. That is, if the corresponding transformed task set can be shown to be schedulable under fixed-priority scheduling using \emph{any} applicable analysis, then the period enforcer algorithm also provides a feasible schedule.

However, note that Theorem 5 in \cite{Raj:suspension1991} applies only to the \emph{transformed} task set: recall that, in the original analysis~\cite{Raj:suspension1991}, deferrable tasks are assumed to defer their entire execution time either completely or not at all (but not parts of it).  Therefore, if we would like to use the period enforcer algorithm to handle segmented self-suspending task sets, we first have to answer the following question: ``\emph{Given a set of sporadic segmented self-suspending tasks, what is the corresponding set of (single-segment) deferrable tasks?}'' That is, how do we convert given self-suspension segments into equivalent bounds on release jitter such that we may apply Theorem 5 to conclude that the system remains schedulable despite period enforcement? Unfortunately, the original proposal \cite{Raj:suspension1991} does not provide an answer to this central question.

In this report, we make two observations pertinent to this question:
\begin{enumerate}
	\item There exist sporadic segmented self-suspending task sets that are schedulable under fixed-priority scheduling without any enforcement, but the corresponding schedule by using the period enforcer algorithm is not feasible. This shows that Theorem 5 in \cite{Raj:suspension1991} has to be  used with care --- it may be applied only in the context of the transformed single-segment deferrable task set, and not in the context of the original segmented self-suspending task set.

@@

\section{Deadline Missses Caused by Period Enforcement}
\label{sec:unschedulable}

In this section, we demonstrate with an example that there exist sporadic  segmented self-suspending task sets that both \textbf{(i)} are schedulable \emph{without} period enforcement and \textbf{(ii)} are not schedulable under the period enforcer.

Consider a task system with $2$ tasks. Let $C_1 = 2$ and $T_1=D_1=10$ (without self-suspension). Let $C_{2,1} = 1, S_{2,1} = 6, C_{2,2}=1, T_2=D_2=11$. Suppose that we use the rate-monotonic priority assignment, i.e., $\tau_1$ has higher priority than $\tau_2$. This task system is feasible without any enforcement since at most one computation segment of a job of $\tau_2$ can be affected by $\tau_1$. The worst-case response time of task $\tau_2$ is hence $10$.

Figure \ref{fig:example} is the resulting schedule by using the period enforcer algorithm. The first job of task $\tau_2$ (that arrives at time $0$) is executed as if there is no period enforcement. The second job (that releases at time $11$) of task $\tau_2$ requires some attention to understand how the period enforcer algorithm works. Note that the first computation segment of task $\tau_2$ does not suffer from any deferrable execution (it is postponed due to higher-priority block). Therefore, it is always placed to the ready queue immediately when it arrives and scheduled as early as possible.

Let's now look at the computation segment $C_2^2$ of the second job of task $\tau_2$. This segment is placed back to the ready queue at time $19$ after its self-suspension. Without period enforcement, this computation segment should be executed from $19$ to $20$. However, the period enforcer algorithm intentionally defers this computation segment to ensure \emph{the periodic demand} of the computation segment $C_2^2$. Therefore, this computation segment is \emph{activated} to the ready queue at time $9+T_2=9+11=20$. This results in a deadline miss.\footnote{By using the notations in \cite{Raj:suspension1991}, we have $ET_{2,1}^2=9, ET_{2,2}^2 = \max\{9+11, 19\}=20$. The superscript $^2$ is for denoting the second computation segment of task $\tau_2$.}

Therefore, the above example shows that there exists (at least) a sporadic segmented self-suspending task system that is schedulable under fixed-priority scheduling without any enforcement, but the corresponding schedule by using the period enforcer algorithm is not feasible. In this example, it is also clear that we can pessimistically convert the two computation segments of task $\tau_2$ into two deferrable tasks $\tau_2^1$ and $\tau_2^2$. The deferrable time of task $\tau_2^1$ is $0$, whereas the deferrable time of task $\tau_2^2$ is at most \emph{$9$}. With the above conversion, we can clearly see that the deferrable task set $\{\tau_1, \tau_2^1, \tau_2^2\}$ is in fact not schedulable by the rate-monotonic priority assignment since we can release a job of $\tau_1$ together with a job of $\tau_2^2$ (after its worst-case deferrable time).

\section{How to Derive the Corresponding Deferrable Task Set}
\label{sec:convert}

The example in Section \ref{sec:unschedulable} can be easily converted to a corresponding deferrable task set. However, it is in general not an easy problem to convert a self-suspending task system into a deferrable task set \emph{precisely}. We can demonstrate the difficulty by looking at a special case. Suppose that the system has $k-1$ ordinary sporadic tasks and only one segmented self-suspending task $\tau_k$. Converting a computation segment into a deferrable task requires to derive the \emph{worst-case resume time of a computation segment}, denoted as $R_k^j$ for the $j$-th computation segment of task $\tau_k$. Suppose that the worst-case response time of the $j$-th computation segment of task $\tau_k$ is $W_k^j$. It is not difficult to see that $R_k^1=0$ and $R_k^j = W_k^{j-1}+S_k^{j-1}$ for $j=2,3,\ldots,m_k-1$. So, we need to derive the worst-case response times of the computation segments of task $\tau_k$.

With the above discussions, for the most simple case defined above, the problem is basically identical to the worst-case response time analysis of segmented self-suspending task systems. It has been recently shown by Nelissen et al. \cite{ecrts15nelissen}  calculating the worst-case response time in the above case is already a very challenging problem. Nelissen et al. \cite{ecrts15nelissen} identify several misconceptions used in the literature for this problem. After excluding those misconceptions, deriving the worst-case response time of a computation segment in pseudo-polynomial time seems to be a very difficult problem. Due to \cite{ecrts15nelissen}, the only existing solution to derive the exact $W_k^{j}$ (and hence $R_k^j$) requires exponential time complexity. And, also recall the example in \ref{sec:unschedulable}. Even if the conversion is done precisely, the deferrable task system can be more pessimistic than the original self-suspending task system with respect to the schedulability.

@@

This report explains how to use the period enforcer algorithm proposed by Rajkumar \cite{Raj:suspension1991} to handle segmented self-suspending real-time tasks. One key assumption in the report in \cite{Raj:suspension1991} is that a deferrable task $\tau_i$ can defer its entire execution time but not parts of it. This creates some mismatches between the original self-suspending task system and the corresponding deferrable task system.  The report in  \cite{Raj:suspension1991} did not explain how to convert a segmented self-suspending task system to a corresponding deferrable task system. We show in this report that this is non-trivial and the main claim of Theorem 5 in \cite{Raj:suspension1991} does not reflect the schedulability of the original self-suspending task system.

Nevertheless, the result in Theorem 5 in \cite{Raj:suspension1991} can be very useful for handling self-suspending task systems if there exist \emph{efficient} schedulability tests of the corresponding deferrable task systems or the period enforcer algorithm. However, such tests do not exist yet and remain as open problems.

>>>>

Theorem 5 in \cite{Raj:suspension1991} provides a very positive result for the analysis of deferrable task sets subject to period enforcement: if the corresponding transformed task set can be shown to be schedulable under fixed-priority scheduling using \emph{any} applicable analysis, then the period enforcer algorithm also yields a correct schedule.

However, note that Theorem 5 in \cite{Raj:suspension1991} applies only to the \emph{transformed} task set: recall that, in the original analysis~\cite{Raj:suspension1991}, deferrable tasks are assumed to defer their entire execution time either completely or not at all (but not parts of it).  Therefore, if we would like to use the period enforcer algorithm to handle segmented self-suspending task sets, we first have to answer the following question: ``\emph{Given a set of sporadic segmented self-suspending tasks, what is the corresponding set of (single-segment) deferrable tasks?}'' That is, how do we convert given self-suspension segments into equivalent bounds on release jitter such that we may apply Theorem 5 to conclude that the system remains schedulable despite period enforcement?

In this report, which is motivated by the fact that the original proposal \cite{Raj:suspension1991} does not provide an answer to this central question, 
we make two pertinent observations.
\begin{enumerate}
	\item There exist sporadic segmented self-suspending task sets that are schedulable under fixed-priority scheduling without any enforcement, but the corresponding schedule by using the period enforcer algorithm is not feasible. This shows that Theorem 5 in \cite{Raj:suspension1991} has to be  used with care --- it may be applied only in the context of the transformed single-segment deferrable task set, and not in the context of the original segmented self-suspending task set.

@@

\section{Period Enforcement Can Induce Deadline Misses}
\label{sec:unschedulable}

In this section, we demonstrate with an example that there exist sporadic  segmented self-suspending task sets that both \textbf{(i)} are schedulable \emph{without} period enforcement and \textbf{(ii)} are not schedulable with period enforcement.

Consider a task system with $2$ tasks. Let $\tau_1$ denote a sporadic task without self-suspensions and parameters $C_1 = 2$ and $T_1=D_1=10$, and let $\tau_2$ denote a self-suspending task consisting of two segments with parameters  $C_{2,1} = 1$,  $S_{2,1} = 6$, $C_{2,2}=1$, and $ T_2=D_2=11$. Suppose that we use the rate-monotonic priority assignment, i.e., $\tau_1$ has higher priority than $\tau_2$. This task set is schedulable without any enforcement since at most one computation segment of a job of $\tau_2$ can be delayed by $\tau_1$: 
\begin{itemize}
	\item if the first segment of a job of $\tau_2$ is interfered with by $\tau_1$, then the second segment resumes at most after $9$ time units after the release of the job and the response time of task $\tau_2$ is hence $10$; otherwise,
	\item  if the first segment of a job of $\tau_2$ is not interfered with by $\tau_1$, then the second segment resumes at most $7$ time units after the release of the job and hence the  response time of task $\tau_2$ is at most $10$ even if the second segment is interfered with by $\tau_1$.
\end{itemize}

\textbf{XXX TODO:} It would be nice to show a schedule for the uncontrolled example.\textbf{XXX}

The period enforcer algorithm determines for each segment (i.e., for each converted single-segment deferrable task) an \emph{eligibility time}. If a segment resumes (i.e., a job of a single-segment deferrable task arrives) before its eligibility time, the execution of the segment (i.e., the single-segment job) is delayed until the eligibility time is reached.

Let $ET_{2,j}^1$ and $ET_{2,j}^2$ denote the eligibility times of the first and second segment, respectively, of the $j^{\mathrm{th}}$ job of $\tau_2$. Further, let $a^1_{2,j}$ and $a^2_{2,j}$ denote the arrival times\footnote{A segment \emph{arrives} when it becomes available for execution. The first segment arrives immediately when the job is released; the segment arrives when the job resumes from its self-suspension.} of the job's first and second segment, respectively.  Finally, let $\mathit{busy}(\tau_2, t')$ denote the last time that a level-2 busy interval began on or prior to time $t'$. According to Section 3.1 in \cite{Raj:suspension1991}, the period enforcer algorithm then defines the segment eligibility times as
\begin{align*}
	ET_{2,j}^1 & = \max\left(ET_{2,j-1}^1 + T_2,\ \mathit{busy}(\tau_2, a^1_{2,j})\right)  \text{ and }
\\
	ET_{2,j}^2 & = \max\left(ET_{2,j-1}^2 + T_2,\ \mathit{busy}(\tau_2, a^2_{2,j}) \right),
\end{align*}
where $ET_{2,0}^1 = ET_{2,0}^2 = -T_2$.

Figure \ref{fig:example} shows the resulting schedule assuming a periodic release pattern. The first job of task $\tau_2$ (which arrives at time $a^1_{2,1} =  0$) is executed as if there is no period enforcement since the definition $ET_{2,0}^1 = ET_{2,0}^2 = -T_2$ ensures that both segments are immediately eligible. Note that the first segment of $\tau_2$'s first job is delayed due to interference from $\tau_1$. As a result, the second segment of $\tau_2$'s first job does not resume until time $a^2_{2,1} = 9$. Thus, we have
\begin{align*}
	ET_{2,1}^1 & = \max\left(-T_2 + T_2,\ \mathit{busy}(\tau_2, 0)\right) = 0  \text{ and }
\\
	ET_{2,1}^2 & = \max\left(-T_2 + T_2,\ \mathit{busy}(\tau_2, 9\right) ) = 9.
\end{align*}

In contrast to the first job, the second job of task $\tau_2$ (which is released at time $11$) is affected by period enforcement. The first segment of the second job arrives at time $a^1_{2,2} = 11$, is interfered with for one time unit, and suspends at time $13$. The segment second of the second job hence resumes only at time $a^2_{2,2} = 19$. Thus, we have
\begin{align*}
	ET_{2,2}^1 & = \max\left(0 + 11,\ \mathit{busy}(\tau_2, 11)\right) = 11  \text{ and }
\\
	ET_{2,2}^2 & = \max\left(9 + 11,\ \mathit{busy}(\tau_2, 19\right) ) = 20.
\end{align*}
According to the rules of the period enforcer algorithm, the processor therefore remains idle at time $19$ because the segment is not eligible to execute until time $ET_{2,2}^2 = 20$. However, at time $20$, the third job of $\tau_1$ is released. As a result, the second job of $\tau_2$ incurs additional interference and misses its deadline at time $22$.

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.9\columnwidth]{figures/example}
\caption{An illustrative example demonstrating a deadline miss at time $22$ under the period enforcer algorithm.
\label{fig:example}%
}
\end{center}
\end{figure}

This example shows that there exist sporadic segmented self-suspending task sets that   \textbf{(i)} are schedulable under fixed-priority scheduling without any enforcement, but \textbf{(ii)} are not schedulable under the period enforcer algorithm.

Furthermore, the example also demonstrates that the conversion to single-segment deferrable tasks does incur a loss of generality since it introduces pessimism: if we convert the two computation segments of task $\tau_2$ into two deferrable tasks $\tau_2^1$ and $\tau_2^2$, where  task $\tau_2^1$ never defers its execution and task $\tau_2^2$ defers its execution by at most \emph{$9$} time units,  then it is immediately obvious that the resulting single-segment deferrable task set $\{\tau_1, \tau_2^1, \tau_2^2\}$ is in fact not schedulable under a rate-monotonic priority assignment since we can time the release of a job of $\tau_1$ to coincide with the arrival of a job of $\tau_2^2$ after it has maximally deferred its execution, which clearly results in a deadline miss.

\clearpage

\section{Deriving a Corresponding Deferrable Task Set}
\label{sec:convert}

The example in Section \ref{sec:unschedulable} can be easily converted to a corresponding single-segment deferrable task set. However, in general,  \emph{precisely} converting a self-suspending task system into a corresponding single-segment  deferrable task set is not an easy problem. We can demonstrate the inherent difficulty by looking at a special case.

Suppose that the system has $k-1$ ordinary sporadic tasks and only one segmented self-suspending task $\tau_k$. Converting a computation segment into a deferrable task requires to derive the \emph{worst-case resume time of a computation segment}, denoted as $R_k^j$ for the $j^{\mathrm{th}}$ computation segment of task $\tau_k$. Suppose that the worst-case response time of the $j^{\mathrm{th}}$ computation segment of task $\tau_k$ is $W_k^j$. It is not difficult to see that $R_k^1=0$ and $R_k^j = W_k^{j-1}+S_k^{j-1}$ for $j=2,3,\ldots,m_k-1$. We therefore need to derive the worst-case response times of the computation segments of task $\tau_k$.

Based on these considerations, it appears that, at least for the simple example,  the problem is basically identical to the worst-case response time analysis of segmented self-suspending task systems.  However, it has been recently shown by Nelissen et al.\ \cite{ecrts15nelissen} that calculating the worst-case response time in the above ``simple'' case is already a very challenging problem. In particular, Nelissen et al.\ \cite{ecrts15nelissen} identified several misconceptions in prior analyses, and after correcting those misconceptions, observed that deriving the worst-case response time of a computation segment in pseudo-polynomial time seems to be a very challenging problem.

In the context of the period enforcer, we consequently observe that the only existing solution to derive the \emph{precise} bound $W_k^{j}$ (and hence $R_k^j$), due to Nelissen et al.\ \cite{ecrts15nelissen},  has exponential time complexity. Furthermore, as demonstrated with the example shown in Figure~\ref{fig:example}, even if the conversion is done precisely, the transformed single-segment deferrable task set can admit more pessimism than the original self-suspending task set with respect to schedulability.

@@

We have revisited the period enforcer algorithm proposed by Rajkumar \cite{Raj:suspension1991} to handle segmented self-suspending real-time tasks. One key assumption in the original report \cite{Raj:suspension1991} is that a deferrable task $\tau_i$ can defer its entire execution time but not parts of it. This creates some mismatches between the original self-suspending task set and the corresponding deferrable task set, which we have demonstrated with an example that shows that Theorem 5 in \cite{Raj:suspension1991} does not reflect the schedulability of the original self-suspending task system.

Furthermore, the original report \cite{Raj:suspension1991} left open the question of how to convert a segmented self-suspending task system to a corresponding deferrable task system. Taking into account recent developments~\cite{ecrts15nelissen}, we have observed that such a task set transformation is non-trivial in the general case.

Nevertheless, Theorem 5 in \cite{Raj:suspension1991} can be useful for handling self-suspending task systems should there exist \emph{efficient} schedulability tests for the corresponding deferrable task systems or the period enforcer algorithm. However, such tests have not been found yet and the development of a precise and efficient schedulability test for self-suspending tasks remains an open problem.
